---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

I retrieved the data.

```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if(!file.exists("actmon.zip")){
    download.file(fileUrl, destfile = "actmon.zip")
}

unzip("actmon.zip")

actMonData <- read.csv("activity.csv")

``` 

Convert "date" factor to Date class

```{r}

actMonData$date <- as.Date(actMonData$date)

``` 

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day
```{r}
dailyStepsByDate <- aggregate(actMonData$steps, by=list(actMonData$date), sum)

``` 
Cleanup column names
```{r}
names(dailyStepsByDate) <- c("Date", "Steps")

``` 

Plot histogram of daily steps
```{r}
png("HistogramOfTotalStepsEachDay.png")
hist(dailyStepsByDate$Steps, main = "Total Number of Steps Taken Each Day", xlab = "Number of Steps")
dev.off()

``` 

## What is the average daily activity pattern?

Calculate mean number of steps per day
```{r}
mean(dailyStepsByDate$Steps,na.rm = TRUE)

``` 
Result is 10766.19

Calculate median number of steps per day
```{r}
median(dailyStepsByDate$Steps,na.rm = TRUE)

``` 
Result is 10765

Calculate the mean number of steps taken by interval
```{r}
dailyStepsByInterval <- tapply(actMonData$steps, actMonData$interval, mean, na.rm = TRUE)

``` 

Plot time series of average daily steps
```{r}
png("TimeSeriesOfAverageStepsEachDay.png")
plot(row.names(dailyStepsByInterval), dailyStepsByInterval, type="l", main = "Average Daily Number of Steps by Interval", xlab = "Interval", ylab = "Average Daily Number of Steps")
dev.off()

``` 

Find interval with maximnum number of steps
```{r}
maxInterval <- which.max(dailyStepsByInterval)
names(maxInterval[1])

``` 
Result is 835

## Imputing missing values

Count NAs
```{r}
sum(is.na(actMonData$steps))

``` 

Result is 2304

Copy Data
```{r}
actMonDataFilledNas <- actMonData

``` 

Replace NAs with average for appropriate interval
```{r}
for (naRowNumber in which(is.na(actMonDataFilledNas$steps))) {
    
    interval <- actMonDataFilledNas[naRowNumber,]$interval
    replacingValue <- dailyStepsByInterval[toString(interval)]
    actMonDataFilledNas[naRowNumber,]$steps <- replacingValue
}


``` 

Calculate the total number of steps taken per day
```{r}
dailyStepsByDateWithNasFilled <- aggregate(actMonDataFilledNas$steps, by=list(actMonDataFilledNas$date), sum)

``` 

Cleanup column names
```{r}
names(dailyStepsByDateWithNasFilled) <- c("Date", "Steps")

``` 

Plot histogram of daily steps
```{r}
png("HistogramOfTotalStepsEachDayWithNasFilled.png")
hist(dailyStepsByDateWithNasFilled$Steps, main = "Total Number of Steps Taken Each Day (NAs Filled)", xlab = "Number of Steps")
dev.off()

``` 

Calculate mean number of steps per day
```{r}
mean(dailyStepsByDateWithNasFilled$Steps,na.rm = TRUE)

``` 

Result is 10766.19, which is the same as prior to "filling in" NAs

Calculate median number of steps per day
```{r}
median(dailyStepsByDateWithNasFilled$Steps,na.rm = TRUE)

``` 
Result is 10766.19, which is greater than prior to "filling in" NAs

## Are there differences in activity patterns between weekdays and weekends?

Add WeekdayWeekend column
```{r}
actMonDataFilledNas$WeekdayWeekend <- factor((weekdays(actMonDataFilledNas$date) %in% c("Saturday", "Sunday")), levels = c(FALSE, TRUE), labels=c("Weekday", "Weekend"))


``` 

Calculate the mean number of steps taken by interval
```{r}
weekdayData <- actMonDataFilledNas[which(actMonDataFilledNas$WeekdayWeekend == "Weekday"),]
weekendData <- actMonDataFilledNas[which(actMonDataFilledNas$WeekdayWeekend == "Weekend"),]

dailyStepsByIntervalWeekdays <- tapply(weekdayData$steps, weekdayData$interval, mean, na.rm = TRUE)
dailyStepsByIntervalWeekends <- tapply(weekendData$steps, weekendData$interval, mean, na.rm = TRUE)


``` 

Plot time series to compare weekend/weekdays
```{r}
png("WeekdayWeekendCompTimeSeries.png")
par(mfrow=c(2,1))
plot(row.names(dailyStepsByIntervalWeekdays), dailyStepsByIntervalWeekdays, type="l", main = "Weekdays", xlab = "Interval", ylab = "Average Daily Number of Steps", ylim= c(0,250))
plot(row.names(dailyStepsByIntervalWeekends), dailyStepsByIntervalWeekends, type="l", main = "Weekends", xlab = "Interval", ylab = "Average Daily Number of Steps", ylim= c(0,250))
dev.off()
``` 
